---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and run rmarkdown::render("README.Rmd") -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
set.seed(0)
options(digits = 4)
```

# distributions3: Probability Distributions as S3 Objects

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/distributions3)](https://CRAN.R-project.org/package=distributions3)
[![Codecov test coverage](https://codecov.io/gh/zeileis/distributions3/branch/main/graph/badge.svg)](https://app.codecov.io/gh/zeileis/distributions3?branch=main)
[![R-CMD-check](https://github.com/zeileis/distributions3/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zeileis/distributions3/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`distributions3` provides a comprehensive and object-oriented toolbox for probability distributions.

**Features:**

- Distributions are provided as S3 objects. Essentially, these are represented as data
  frames of parameters but with a dedicated class, e.g., `"Normal"`, inheriting from `"distribution"`.
   
- Methods `pdf()` (probability density function or probability mass function), `cdf()`
  (cumulative distribution function), `quantile()` (quantile function, inverse of `cdf()`), and
  `random()` (random samples) replace the d/p/q/r functions provided in base R and many other
  packages such as `dnorm()`, `pnorm()`, `qnorm()`, `rnorm()` for the normal distribution.
   
- Methods for computing central moments via `mean()`, `variance()`, `skewness()`, and `kurtosis()`.

- Methods for extracting the `score()` (first derivative of the log-likelihood with respect to the 
  parameters) and `hessian()` (corresponding second derivative.
  
- Numerical fallback implementations for most of these methods in case no dedicated method is
  available for a certain distribution.


**Goals:**

- User-friendly interface for illustrations in introductory statistics classes and also
  advanced probabilistic modeling and regression courses.

- Developer-friendly interface for fitting and assessing probabilistic models, e.g., via
  [gamlss2](https://gamlss-dev.github.io/gamlss2/) and
  [topmodels](https://zeileis.codeberg.page/topmodels/).


## Installation

The stable version of `distributions3` is available from
[CRAN](https://CRAN.R-project.org/package=distributions3):

``` r
install.packages("distributions3")
```

The latest development version can be installed from
[R-universe](https://zeileis.R-universe.dev/distributions3):

``` r
install.packages("distributions3", repos = "https://zeileis.R-universe.dev")
```


## Basic usage

The basic usage of `distributions3` looks like:

```{r}
library("distributions3")

## Normal "IQ" distribution
X <- Normal(100, 15)
X

random(X, 5)

pdf(X, 85)
cdf(X, 85)
quantile(X, 0.159)

mean(X)
variance(X)
skewness(X)

## Vector of Poisson "soccer goals" distributions
Y <- Poisson(c(0.8, 1.2, 1.9, 0.5))
Y

random(Y, 5)

pdf(Y, 0:5)
cdf(Y, 0:5)
quantile(Y, 0.5)

mean(Y)
variance(Y)
skewness(Y)
```


## Contributing

If you are interested in contributing to `distributions3`, please reach out on Github! We are happy to review PRs contributing bug fixes.

Please note that `distributions3` is released with a
[Contributor Code of Conduct](https://zeileis.github.io/distributions3/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Related work

For a comprehensive overview of the many packages providing various distribution related functionality see the [CRAN Task View](https://CRAN.R-project.org/view=Distributions).

- [`distributional`](https://CRAN.R-project.org/package=distributional) provides distribution objects as S3 `vctr` objects.
- [`distr6`](https://CRAN.R-project.org/package=distr6) builds on `distr`, but uses R6 objects
- [`distr`](https://CRAN.R-project.org/package=distr) is similar in spirit to `distributions3`, but is not vectorized, uses S4 objects, and is less focused on documentation. 
- [`fitdistrplus`](https://CRAN.R-project.org/package=fitdistrplus) provides extensive functionality for fitting various distributions but does not treat distributions themselves as objects
